Distribution-oblivious Programming with Dish

### 2019-11-22 pldi/draft.tex
Notes on distributability

\newcommand{\eg}{{\em e.g.}, }
\newcommand{\ie}{{\em i.e.}, }
\newcommand{\etc}{{\em etc.}\xspace}
\newcommand{\vs}{{\em vs.} }
\newcommand{\cmpn}{compartmentalization} 
\newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
\newcommand{\ttt}[1]{\texttt{\small #1}}
\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
\newcommand{\spol}[1]{\scriptsize{\sc#1}}
\newcommand{\pol}[1]{\texttt{\small {\color{purple}#1}}}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\wka}{\ttt{a\textsubscript{1}}}
\newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}

@@

To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation\kk{I am not sure whether stream processing provides information for distributing a computation on its own. Maybe we can say that the shell exposes operator level parallelism, and that this is the first step to distributing a computation}.
\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies \todo{maximal} sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams \kk{Isn't uniq pure? I am not sure how we can handle it}, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.
% TODO: \kk{We have to mention character vs line based parallelism somewhere, and that we choose to see lines as the data quantum}

Commands can be tuned by two mechanisms, namely command-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  their parsing and evaluation is left entirely up to each command \kk{A minor point: the shell does the expansion, and then hands it of to the command (I am saying that because evaluation might be confounded with expansion so we might want to clarify that)}.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

@@

As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
Intuitively,
 (i) the parallel-composition operator exposes \todo{task} parallelism \kk{I have named that operator level parallelism, which one do we prefer?},
 (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
 (iii) the pipe operator enables both task and data parallelism, and \kk{How does it enable data parallelism?}
 (iv)  other composition operators can be viewed as barrier candidates.
To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.

@@

Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying \todo{maximal} distributable sub-expressions.
It converts the AST to a dataflow graph, and runs...
This analysis is not trivial, as many of the shell's dynamic features have to be taken into consideration---examples include shell expansion, environment variables, file input sizes, \etc

\kk{On the dataflow graph, Dish iteratively performs graph transformations that expose data parallelism as task parallelism while preserving correctness.}

Finally, given a distributed pipeline with runtime \todo{annotations}\kk{the runtime just gets a dataflow graph}, \sys needs to schedule and execute it.
This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.

Second, \sys needs to fill-in planing details left blank by the rewriting pass.
This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available. \kk{I am not exactly sure I understand that :) }

@@

\section{Distributability Classes}

\subsection{Comnands}
\label{bg:cmd}

% Sets vs. Classes

\begin{table}[h]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Distributability Classes}.
    Broadly, \unix commands can be broken down into to five classses.
  }
}
\begin{tabular}{l @{\extracolsep{\fill}} lll}
\toprule
Class                           & Example Commands        & POSIX            & Core-utils\\
\midrule
Stateless                       &                         &                  &           \\
Pure                            &                         &                  &           \\
(Distr.) File-System            &                         &                  &           \\
Side-effectful                  &                         &                  &           \\
Irreversibly Side-effectful     &                         &                  &           \\
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular}
\label{tab:classes}
\end{table}

Define distributability 
Explain distributability classes, based on how

\heading{Preliminaries}

\heading{POSIX, GNU Core-utils, and beyond}

\heading{Stateless}

\heading{Pure}

\heading{File-system}

\heading{Side-effectful}

\heading{Irreversably Side-effectful}

@@

\section{Dish Overview}

@@

\kk{Rename this section to panner activators}

@@

\kk{On our dataflow graph model: Our data flow graph model is
  different than most distributed batch and stream processing systems,
  since the shell is a hybrid between the two. Batch processing
  systems have no notion of data order, and stream processing systems
  usually have order, but none of these systems supports this notion
  of reading from different sources in sequence. Our dataflow graph
  model supports this by including information about incoming edge
  order. Another benefit of our dataflow model, is that together with
  the equations that hold on stateless and pure commands, data
  parallelism is exposed in the graph as task parallelism. This allows
  the unoptimized and optimized dataflow graph to have the same
  unified representation, and doesn't lead to an ad-hoc optimization
  of the dataflow graph as done in the other systems (the last bit is
  too aggressive). }

@@

\heading{DSLs for Distribution}
At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} \kk{JetStream has a very different focus} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
simplify certain patterns, 
but do not offer the flexibility of a full-fledged environment.
% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.

>>>>

\newcommand{\eg}{{\em e.g.}, }
\newcommand{\ie}{{\em i.e.}, }
\newcommand{\etc}{{\em etc.}\xspace}
\newcommand{\vs}{{\em vs.} }
\newcommand{\cmpn}{compartmentalization} 
\newcommand{\heading}[1]{\vspace{4pt}\noindent\textbf{#1}\enspace}
\newcommand{\ttt}[1]{\texttt{\small #1}}
\newcommand{\ttiny}[1]{\texttt{\scriptsize #1}}
\newcommand{\tti}[1]{\texttt{\scriptsize #1}}
\newcommand{\spol}[1]{\scriptsize{\sc#1}}
\newcommand{\pol}[1]{\texttt{\small {\color{purple}#1}}}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\wka}{\ttt{a\textsubscript{1}}}
\newcommand{\wkq}{\ttt{q\textsubscript{1-4}}}

\newcommand{\cn}[1]{\mbox{\textcircled{\footnotesize #1}}}
\newcommand{\tcn}[1]{\mbox{\textcircled{\scriptsize #1}}}

\newcommand{\pur}{\cn{\textsc{P}}}
\newcommand{\sta}{\cn{\textsc{S}}}
\newcommand{\dfs}{\cn{\textsc{F}}}
\newcommand{\sid}{\cn{\textsc{E}}}
\newcommand{\irr}{\cn{\textsc{I}}}

\newcommand{\tpur}{\tcn{\textsc{P}}}
\newcommand{\tsta}{\tcn{\textsc{S}}}
\newcommand{\tdfs}{\tcn{\textsc{F}}}
\newcommand{\tsid}{\tcn{\textsc{E}}}
\newcommand{\tirr}{\tcn{\textsc{I}}}

@@

To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.

@@

% A pipeline is a mechanism for program composition that works by chaining together programs (or, \emph{commands}) using pipes.
\unix provides several ways for succinct program composition.
Central among them is the \emph{pipe}, a primitive that passes the output of one program as input to the next.
These programs produce output and consume input concurrently, and possibly at different rates, with the \unix kernel facilitating program scheduling and communication synchronization behind the scenes.
Intermediate programs, termed commands, can also be developed in different programming languages, as long as they conform to a common interface.
\begin{wrapfigure}{r}{0.06\textwidth}
  \vspace{-12pt}
  % \begin{center}
    % \includegraphics[width=0.48\textwidth]{gull}
    \includegraphics[width=0.06\textwidth]{./figs/dish_ex1.pdf}
  % \end{center}
  \vspace{-30pt}
\end{wrapfigure}
This interface exposes a stream of contiguous character lines.
A few characters are special---for example, the newline (\textsc{NL}) character delineates an element of the stream, and the end-of-file (\textsc{EOF}) character signals the end of a stream.
As part of this interface, each command has access to (any combination of) three \emph{standard} streams---input, output, and error.
% the input stream (stdin), the output stream (stdout), and the generating errors or diagnostics to the standard error stream (stderr)
% generating errors or diagnostics to the standard 
% Leveraging this interface, pipelines chain together commands by their standard streams, such that the output stream of one command (stdout) is passed directly as input (stdin) to the next one.

Commands can be tuned by two mechanisms, namely command-line options and environment variables.
Options control the execution of a program---\eg changing the \ttt{sort}ing order.
The shell does not have any visibility into these options; 
  their parsing and evaluation is left entirely up to each command.
Environment variables carry more general information about the surrounding system---\eg file search paths, system defaults, parsing switches \etc 
They are organized in a map from names to values, both of which are strings, and can be evaluated at any point in the program---including as commands, flags, a shell feature termed parameter \emph{expansion}.

@@

As outlined earlier, the key insight behind \sys is that the shell already describes parallel computations over streams.
Intuitively,
 (i) the parallel-composition operator exposes task parallelism,
 (ii) commands such as \ttt{xargs} and \ttt{tee} expose data parallelism,
 (iii) the pipe operator enables both task and data parallelism, and
 (iv)  other composition operators can be viewed as barrier candidates.
To leverage these insights and correctly distribute a pipeline such as $p_1$, \sys must solve several challenges.

@@

Given a set of commands composed together by the shell primitives, \sys's next challenge is to analyze it and identify candidate subexpressions for distribution.
\sys addresses this challenge with a rewriting \todo{pass} that performs dataflow analysis identifying maximal distributability sub-expressions.
It converts the AST to a dataflow graph, and runs...
This analysis is not trivial, as many of the shell's dynamic features have to be taken into consideration---examples include shell expansion, environment variables, file input sizes, \etc

Finally, given a distributed pipeline with runtime \todo{annotations}, \sys needs to schedule and execute it.
This amounts to solving a few challenges, achieved by \sys's runtime component, two of which stand out.
First, \sys needs to wrap single-node commands with corrective runtime interposition based on their classes;
  this is achieved by thin wrappers for all the commands involved in the program, generated just-in-time using the distributability specification outlined earlier.
The reasoning behind wrappers being thin is to not interfere with legacy functionality---commands have many flags accumulated over long periods of time.

Second, \sys needs to fill-in planing details left blank by the rewriting pass.
This is achieved with \emph{planner activators}, commands that yield (and pass their arguments) to the scheduler.
From the point of view of the shell, activators are higher-order commands \'{a} la \ttt{xargs} and \ttt{time} that take other commands as arguments---a technique that achieves the latest possible binding, when critical runtime information is available.

@@

\section{Distributability Classes}
\label{distributability}

To have any hope of distributing \unix pipelines, \sys first needs to understand the distribution characteristics of individual commands involved in them.
This understanding is encoded in a command's \emph{distributability} class.
Put informally, a distributability class representing the level of synchronization required by copies of a command executing concurrently.
The class captures characteristics that are imprortant for distribution, rather than all the observable behavior of a command.
\sys leans towards a few, coarse classes rather than many, detailed ones---after all, these classes should be easily understood by command developers.
% Make the connections that this is like a type, that doesn't capture everything, but is good-enough

The section starts with a distributability study of standard libraries commands, such as {\sc POSIX} and GNU Coreutils~\sx{cmd}.
Building on this study, it presents a DSL for describing a command's distributability characteristics~\sx{ext}.
In return, \sys uses this DSL to tag commands in the standard libraries and generate the transformations presented in later sections.

\subsection{Distributability of Standard Libraries}
\label{cmd}

% nv TODO: this is a hierarchy
Broadly, shell commands can be split into five major classes with respect to their distribution characteristics, depending on how they interact with state.
Tab.~\ref{tab:classes} summarizes the discussion;
These classes are listed (and presented later) in ascending difficulty of distribution; later classes require further effort.
In this order, each class can be thought as a subset of the next---\eg all stateless commands are pure---meaning that the synchronization mechanisms required for any superclass would work with its subclass (without seeing any performance improvements).

\begin{table}[t]
\center
\footnotesize
\setlength\tabcolsep{3pt}
\caption{
  \footnotesize{
    \textbf{Distributability Classes}.
    Broadly, \unix commands can be broken down into to five classes.
  }
}
\begin{tabular}{l @{\extracolsep{\fill}} llll}
\toprule
Class                           &  Key    & Example Commands                            & Coreutils       & POSIX       \\
\midrule
Stateless                       & ~\tsta  & \tti{tr},   \tti{cat},    \tti{grep}        &  22 (21.1\%)     &           \\  % 22
Pure                            & ~\tpur  & \tti{sort}, \tti{wc},     \tti{uniq}        &  21 (20.1\%)     &           \\  % 21
(Distr.) File-System~           & ~\tdfs  & \tti{cp},   \tti{ln},     \tti{rm}          &  27 (25.9\%)     &           \\  % 27
Side-effectful                  & ~\tsid  & \tti{env},  \tti{chroot}, \tti{whoami}      &  25 (24\%)       &           \\  % 25
Irreversible                    & ~\tirr  & \tti{exit}, \tti{lpr},    \tti{restart}     &  5  (0.4\%)      &           \\  % 5
% \midrule
% Shell PL Constructs             &                         &                  \\
% \etc                            &                         &                  \\
\bottomrule
\end{tabular}
\label{tab:classes}
\end{table}

% \heading{Preliminaries}
% \heading{POSIX, GNU Core-utils, and beyond}

\heading{Stateless Commands}
The first class contains stateless commands~\sta that operate on individual elements of the stream, without maintaining state across invocations.
These are commands expressible as a purely functional map---\eg \ttt{basename} removes a path prefix from a string and \ttt{grep} filters out individual lines.
At times, stateless commands may operate as functions that produce multiple elements---\eg when they insert a {\sc NL} token.
Workloads that use only stateless commands are trivial to parallelize:
  they do not require any synchronization to maintain correctness,
  they do not need care planning on where to split.
 % map :: (a -> b) -> [a] -> [b]

Many of these commands (about 1/3 of coreutils' \sta) are stateless \emph{within} an element (\eg \ttt{tr} transliterates characters, one at a time);
  this could potentially allow further parallelization within a single line element.
However, exactly because of their local, commands operating on individual characters are not particularly expensive computationally.

\heading{Pure Commands}
The second class contains pure commands \pur.
These commands respect functional purity, returning the same results for the same inputs;
  however, they need to maintain internal state across the entire invocation.

Pure commands often cannot operate in an online fashion, but need to block until the end of the stream.
A typical example of this is \ttt{sort}, a command that cannot start before the last line of input has been consumed.

Parallelizing these commands is slightly more challenging and depends on the command.
Some commands have known divide-and-conquer equivalents.

Pure commands~ones that maintain 
Distribution:

\heading{File-system Commands}

This is the largest class, as the file (and file-system) abstraction is so central to \unix.
In fact, \unix proxies several file-unrelated operators through the file-system---for example, security in \unix is grafted upon the file-system:

(and all associated commands) are always
And most of these commands seem control-plane operations that operate on identifiers~\sx{bg}---for example, renaming a file or changing group ownership.
% TODO: mention that some commands fail to comply with the interface---for example, csplit could be split followed by a redirection to files.
These are mostly distributable

Distribution:

\heading{Side-effectful Commands}
Another large class of commands are ones that have side-effects to memory.
Because of timesharing and multi-tasking (many users can operate multiple terminals on single \unix machine), a set of commands are devoted to accessing variables---environment variables, shell parsing flags, 
The different formats have multiplied these commands

Thankfully, the vast majority of these commands (a ratio of 21:4) only \emph{read} values;
Even better,  usually read values that are not altered by user---\eg \ttt{date}, \ttt{uname}, and \ttt{finger} are all commands that interface with kernel- or hardware-generated information.
Even if \sid commands are protected with transactions
these benefits are even greater for read-only or read-mostly accesses.

\heading{Irreversibly Side-effectful Commands}

@@

@@

@@

@@

\heading{DSLs for Distribution}
At the other end of the spectrum, distributed computing frameworks~\cite{mapreduce:08, ciel:11, spark:12, naiad:13, jetstream:14} and domain-specific languages~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp, psync:16, dartagnan:18}
simplify certain patterns, 
but do not offer the flexibility of a full-fledged environment.
% suffer from a similar lack of generality by making strong assumptions about the nature of the computation---for example, strongly-eventual commutative components that can proceed in parallel.
Developing under these frameworks differs quite significantly from the development of normal (non-distributed) programs.
